---
title: "Getting started with hdstats"
author: "Yikai Zhang, Qian Tang, and Boxiang Wang"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started with hdstats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4.5
)
```

# Overview

The **hdstats** package fits the whole regularization path of four
penalized regression and classification models for high-dimensional data,
all sharing one interface and one computational engine:

| Function    | Loss                         | Response            | Model-specific argument |
|-------------|------------------------------|---------------------|-------------------------|
| `hdhuber()` | Huber                        | continuous          | `delta` (transition point) |
| `hdqr()`    | quantile (check) loss        | continuous          | `tau` (quantile level) |
| `hdrr()`    | Wilcoxon rank (Jaeckel)      | continuous          | none (built on `hdqr()`) |
| `hdsvm()`   | hinge (support vector machine) | binary            | none |

Every model minimizes, over the intercept $b_0$ and the coefficients
$\beta \in \mathbb{R}^p$,

$$
\frac{1}{n}\sum_{i=1}^n \ell(y_i, b_0 + x_i^\top\beta)
+ \lambda_1 \sum_{j=1}^p w_j |\beta_j|
+ \frac{\lambda_2}{2} \sum_{j=1}^p v_j \beta_j^2 ,
$$

where $\ell$ is the model's loss, $\lambda_1$ (argument `lambda`) is the
lasso tuning parameter for which a whole path of solutions is computed,
$\lambda_2$ (argument `lam2`) is a single ridge parameter, and $w_j$ and $v_j$
(arguments `pf` and `pf2`) are optional penalty factors that allow adaptive
penalties or unpenalized variables.

The Huber loss is smooth and is minimized directly by coordinate descent.
The hinge and check losses are not differentiable; they are handled by the
finite smoothing algorithm of Tang, Zhang, and Wang (2024), which replaces
the loss by its convolution with a uniform kernel of bandwidth `hval`, solves
the smoothed problem by coordinate descent with warm starts and active-set
updates, and then shrinks the bandwidth and refines the solution until the
KKT conditions of the original problem are met. Nonconvex SCAD and MCP
penalties are available for the SVM and quantile models through the local
linear approximation (`nc.hdsvm()`, `nc.hdqr()`). Every model has a `cv.*()`
function for $k$-fold cross-validation, and all objects have `coef()`,
`predict()`, and `print()` methods; cross-validation objects also have a
`plot()` method.

# Simulated data

We simulate a sparse regression problem with $n = 100$ observations,
$p = 200$ predictors, and heavy-tailed $t_3$ errors, a setting in which
robust losses are attractive.

```{r data}
library(hdstats)
set.seed(2024)
n <- 100
p <- 200
x <- matrix(rnorm(n * p), n, p)
beta_star <- c(2, -1.5, 1, 0.8, rep(0, p - 4))
y <- drop(x %*% beta_star) + rt(n, df = 3)
```

# Huber regression

`hdhuber()` computes the solution path for a decreasing sequence of
`lambda` values (100 by default). Printing a fit shows the number of nonzero
coefficients at each `lambda`.

```{r huber}
fit <- hdhuber(x, y, delta = 1, lam2 = 0.01)
fit$lambda[1:5]
fit$df[1:10]
# print() lists Df and lambda along the path; shown here for a short path
hdhuber(x, y, delta = 1, lam2 = 0.01, nlambda = 10)
```

`coef()` and `predict()` accept any value of `lambda` through the argument
`s`; values that are not on the fitted grid are handled by linear
interpolation.

```{r huber-coef}
coef(fit, s = 0.1)[1:6, ]
predict(fit, newx = x[1:3, ], s = c(0.2, 0.1))
```

`cv.hdhuber()` performs 5-fold cross-validation (by default) using the Huber
loss on the held-out folds, and `plot()` shows the cross-validation curve
with one-standard-error bars; the dotted lines mark `lambda.min` and
`lambda.1se`. All `cv.*()` functions accept `ncores` to fit the folds in
parallel.

```{r huber-cv}
cv.fit <- cv.hdhuber(x, y, delta = 1, lam2 = 0.01)
cv.fit
plot(cv.fit)
```

Coefficients at the selected tuning parameter:

```{r huber-cv-coef}
b <- coef(cv.fit, s = "lambda.min")
b[b[, 1] != 0, , drop = FALSE]
```

# Quantile regression

`hdqr()` has the same interface with the quantile level `tau`
(default 0.5, i.e. median regression). Different quantiles can reveal
heteroscedastic effects.

```{r qr}
fit.med <- hdqr(x, y, tau = 0.5, lam2 = 0.01)
fit.q90 <- hdqr(x, y, tau = 0.9, lam2 = 0.01)
cbind(median = coef(fit.med, s = 0.1)[1:5, ],
      q90 = coef(fit.q90, s = 0.1)[1:5, ])
```

Cross-validation uses the check loss at the same quantile level:

```{r qr-cv}
cv.q <- cv.hdqr(x, y, tau = 0.5, lam2 = 0.01)
plot(cv.q)
```

## Nonconvex penalties

The lasso penalty shrinks large coefficients towards zero. The SCAD and MCP
penalties reduce this bias; `nc.hdqr()` fits them with a few local linear
approximation steps, each of which is a weighted lasso fit. The
`lambda` sequence for the nonconvex penalty can be supplied or, by default,
taken from the initial lasso fit; here we use a short grid to keep the
computation small.

```{r nc-qr}
lambda <- 10^seq(-0.5, -1.5, length.out = 10)
nc.fit <- nc.hdqr(x, y, tau = 0.5, lambda = lambda, lam2 = 0.01, pen = "scad")
cbind(lasso = coef(fit.med, s = 0.1)[1:5, ],
      scad = coef(nc.fit, s = 0.1)[1:5, ])
```

`cv.nc.hdqr()` cross-validates the nonconvex fit in the same way.

```{r nc-qr-cv}
cv.nc <- cv.nc.hdqr(x, y, tau = 0.5, lambda = lambda, lam2 = 0.01, pen = "mcp")
cv.nc$lambda.min
```

# Wilcoxon rank regression

Jaeckel's rank regression with Wilcoxon scores minimizes the dispersion of
the residuals, $\sum_{i<j} |r_i - r_j|$, which is the median check loss of all
pairwise differences of the data. `hdrr()` builds the $n(n-1)/2$ differences
and calls `hdqr()`, then recovers the intercept as the median residual. The
returned object inherits the `hdqr` methods.

```{r rr}
rr.fit <- hdrr(x, y, lam2 = 0.01, nlambda = 50)
coef(rr.fit, s = 0.1)[1:5, ]
cv.rr <- cv.hdrr(x, y, lam2 = 0.01, nlambda = 50)
cv.rr
```

Because the design is differenced, memory and time grow as $O(n^2 p)$, so
`hdrr()` is intended for moderate $n$.

# Support vector machines

`hdsvm()` fits the elastic-net penalized linear SVM to a binary response,
which may be coded as $-1/1$, $0/1$, or a two-level factor. Predictions are
class labels by default or the linear predictor with `type = "loss"`.

```{r svm}
prob <- plogis(drop(x %*% beta_star))
yb <- factor(ifelse(rbinom(n, 1, prob) == 1, "case", "control"),
             levels = c("control", "case"))
svm.fit <- hdsvm(x, yb, lam2 = 0.01)
cv.svm <- cv.hdsvm(x, yb, lam2 = 0.01)
plot(cv.svm)
table(predicted = predict(cv.svm, x, s = "lambda.min"), observed = yb)
```

The nonconvex version works in the same way:

```{r nc-svm}
nc.svm <- nc.hdsvm(x, yb, lambda = lambda, lam2 = 0.01, pen = "scad")
nc.svm$df
```

# Common arguments

* `lambda`, `nlambda`, `lambda.factor`: the `lambda` sequence is computed
  automatically from `lambda.factor * lambda_max` to `lambda_max` on a log
  scale, or can be supplied (it is sorted in decreasing order so that warm
  starts can be used).
* `lam2`: the ridge parameter $\lambda_2$; a single value per fit.
* `pf`, `pf2`, `exclude`: L1 and L2 penalty factors (0 means unpenalized)
  and variables to drop; `pf` can also be a matrix with one column per
  `lambda`, which is how the nonconvex fits are implemented.
* `standardize`: predictors are standardized internally by default;
  coefficients are always returned on the original scale.
* `dfmax`, `pmax`: stop the path early once too many variables are active.
* `eps`, `maxit`: convergence threshold and the maximum number of
  coordinate-descent passes for the whole path; a warning is issued and the
  partial path is returned if `maxit` is reached.
* `hval`, `is_exact`, `sigma` (`hdqr()`, `hdsvm()`): the initial smoothing
  bandwidth, whether to run an additional exact projection step at small
  bandwidths, and the augmented-Lagrangian parameter of that step.

# References

Tang, Q., Zhang, Y., and Wang, B. (2024). Finite smoothing algorithm for
high-dimensional support vector machines and quantile regression.
*Proceedings of the 41st International Conference on Machine Learning*.
<https://openreview.net/forum?id=RvwMTDYTOb>

Friedman, J., Hastie, T., and Tibshirani, R. (2010). Regularization paths
for generalized linear models via coordinate descent. *Journal of
Statistical Software*, 33(1), 1--22.
<https://doi.org/10.18637/jss.v033.i01>
